Discrete Probability for Discrete Quantum Computing

### 2016-05-04 prob.tex
mess

In this case, we split the $[0,1]$ interval into a collection of
overlapping closed sub-intervals obeying certain restrictions to be
described at the end of the section. First we illustrate the main
ideas using a simple example.

\begin{example}[Two-coin probability space with four intervals] \label{ex3}
We split the unit interval $[0,1]$ in the following four closed sub-intervals:
\begin{itemize}
\item $[0,0]$ which we call \imposs, 
\item $[0,\frac{1}{2}]$ which we call \unlikely, 
\item $[\frac{1}{2},1]$ which we call \likely, and 
\item $[1,1]$ which we call \necess.
\end{itemize}
Using these new values, the probability measure becomes:
\[
\begin{array}{c@{\qquad\qquad}c}
\begin{array}{rcl}
\pmeas(\emptyset) & = & \imposs\\
\pmeas(\{HH\}) & = & \unlikely\\
\pmeas(\{HT\}) & = & \imposs\\
\pmeas(\{TH\}) & = & \likely\\
\pmeas(\{TT\}) & = & \imposs\\
\pmeas(\{HH,HT\}) & = & \unlikely\\
\pmeas(\{HH,TH\}) & = & \necess\\
\pmeas(\{HH,TT\}) & = & \unlikely
\end{array} & \begin{array}{rcl}
\pmeas(\{HT,TH\}) & = & \likely\\
\pmeas(\{HT,TT\}) & = & \imposs\\
\pmeas(\{TH,TT\}) & = & \likely\\
\pmeas(\{HH,HT,TH\}) & = & \necess\\
\pmeas(\{HH,HT,TT\}) & = & \unlikely\\
\pmeas(\{HH,TH,TT\}) & = & \necess\\
\pmeas(\{HT,TH,TT\}) & = & \likely\\
\pmeas(\{HH,HT,TH,TT\}) & = & \necess
\end{array}\end{array}
\]
Not only does this probability measure reveal that the second coin is
double-headed but it also reveals the bias in the first coin, again despite
the absence of any numeric information.  
\qed\end{example}

The probability measure above can be understood in reference to
Ex.~\ref{ex1} by mapping each numeric value to the smallest interval
containing it. The probability measure appears quite intuitive from
that perspective. However, we also must be able to understand the
probability measure independently of Ex.~\ref{ex1} and in particular
justify how combinations such as
\[ 
\pmeas(\{HH\}) = \unlikely, \quad\pmeas(\{TH\}) = \likely, \quad
\pmeas(\{HH\}\cup\{TH\}) = \necess 
\]
which assert that an \unlikely-event whose probability is in the range
$[0,\frac{1}{2}]$ and a \likely-event whose probability is in the
range $[\frac{1}{2},1]$ combine to a \necess-event whose probability
is in the range $[1,1]$! To understand this mystery, we look at
another more general example and then give the formal mathematical
definition of interval-valued probabilities.

\amr{See A Simple View of the Dempster-Shafer Theory of Evidence and 
  its Implication for the Rule of Combination Lotfi A. Zadeh, AAAI,
  1986}

>>>>

A natural generalization of the disjoint set-valued measure above is
to allow the sets to overlap. In this case, we split the interval
$[0,1]$ in a collection of \emph{overlapping} closed
sub-intervals. First we illustrate the main ideas using a simple
example.

\begin{example}[Two-coin probability space with four
  intervals] \label{ex3} We split the unit interval $[0,1]$ in the
  following four closed sub-intervals: $[0,0]$ which we call \imposs,
  $[0,\frac{1}{2}]$ which we call \unlikely, $[\frac{1}{2},1]$ which
  we call \likely, and $[1,1]$ which we call \necess.  Using these new
  values, we can modify the probability measure of Ex.~\ref{ex1} by
  mapping each numeric value to the smallest sub-interval containing
  it to get the following:
\[
\begin{array}{c@{\qquad\qquad}c}
\begin{array}{rcl}
\pmeas(\emptyset) & = & \imposs\\
\pmeas(\{HH\}) & = & \unlikely\\
\pmeas(\{HT\}) & = & \imposs\\
\pmeas(\{TH\}) & = & \likely\\
\pmeas(\{TT\}) & = & \imposs\\
\pmeas(\{HH,HT\}) & = & \unlikely\\
\pmeas(\{HH,TH\}) & = & \necess\\
\pmeas(\{HH,TT\}) & = & \unlikely
\end{array} & \begin{array}{rcl}
\pmeas(\{HT,TH\}) & = & \likely\\
\pmeas(\{HT,TT\}) & = & \imposs\\
\pmeas(\{TH,TT\}) & = & \likely\\
\pmeas(\{HH,HT,TH\}) & = & \necess\\
\pmeas(\{HH,HT,TT\}) & = & \unlikely\\
\pmeas(\{HH,TH,TT\}) & = & \necess\\
\pmeas(\{HT,TH,TT\}) & = & \likely\\
\pmeas(\{HH,HT,TH,TT\}) & = & \necess
\end{array}\end{array}
\]
This probability measure is more informative than the one in
Ex.~\ref{ex2}: not only does it reveal that the second coin is
double-headed but it also reveals the bias in the first coin.
\qed\end{example}

The probability measure above appears quite intuitive but it is not
really evident that it is well-defined. For example, how do we justify
the following combination of assignments:
\[ 
\pmeas(\{HH\}) = \unlikely, \quad\pmeas(\{TH\}) = \likely, \quad
\pmeas(\{HH\}\cup\{TH\}) = \necess 
\]
which assert that an \unlikely-event whose probability is in the range
$[0,\frac{1}{2}]$ and a \likely-event whose probability is in the
range $[\frac{1}{2},1]$ combine to a \necess-event whose probability
is in the range $[1,1]$. To understand this mystery, we look at
another more general example and then give the formal mathematical
definition of interval-valued probabilities.

\begin{example}[Dempster-Shafer Theory of Evidence] We have three
  employees whose precise ages $A_1$, $A_2$, and $A_3$ are not
  known. All is given is a range of ages for each employee:
  $A_1 \in \{ 23,24 \}$, $A_2 \in \{ 20,21,22\}$, and
  $A_3 \in \{ 21,22 \}$. The sample space in this case is
  $A_1 \times A_2 \times A_3$ which represents all the possible
  combinations of ages for the three employees:
\[\begin{array}{rcl}
\Omega &=& \{ 
        (23,20,21), (23,20,22), (23,21,21), (23,21,22), (23,22,21), (23,22,22), \\
&& ~(24,20,21), (24,20,22), (24,21,21), (24,21,22), (24,22,21), (24,22,22) \}
\end{array}\]
Subsets of $\Omega$ represent events as usual. Consider the event
$\Omega$ that \emph{some} employee's age is in the range
$\{20,21,22\}$: since that event covers the entire sample space its
probability must be 1. We can however produce a more informative
answer by reasoning as follows: it is \emph{impossible} for the first
employee's age to be in the range $\{20,21,22\}$; it is \emph{certain}
that the second employee's age is in that range; and it is
\emph{possible} that the third's employee age is in that
range. Aggregating the results, we see that it is \emph{necessary} for
one out of three employees to have their age in the required range,
and it is \emph{possible} for an additional employee to have their age
in the required range. We summarize this information by reporting that
the probability of this event is $[\frac{1}{3},\frac{2}{3}]$ where the
first number reports the \emph{certainty} of the event and the second
reports the \emph{possibility} of the event. We could also have
reasoned about the dual event that \emph{no} employee's age is in the
range $\{20,21,22\}$ to get the probability $[\frac{1}{3},\frac{2}{3}]$

 dually
about the evidence \emph{against} the event. The probability in this
case

Now consider the event
that some employee's age is in the range $\{23,24\}$. Reasoning as
above, the probability of this event is $[\frac{1}{3},\frac{1}{3}]$ as
only the first employee qualifies. Clearly if we were to ask the
probability that some employee's age is in the range
$\{20,21,22,23,24\}$ the result should be $[1,1]$ as this range covers
all the possibilities. The way to calculate this result from the two
previous ones is as follows: the evidence for the combined event to be
necessary is at least as strong as the evidence that each disjoint
event that contributes to it is necessary. So the lower bound
probability for the combined event is at least
$\frac{1}{3}$. Similarly the upper bound probability is at least
$1$. But now we can reason using the complements of the events about
the necessity and possibility of refuting each event. For the event
that some employee's age is \emph{not} in the range $\{20,21,22\}$ we
have a probability $[\frac{1}{3},\frac{1}{3}]$ because it is necessary
that $A_1$ is not in that range. Similary, the event that some employee's age is
\emph{not} in the range $\{23,24\}$ is
$[\frac{2}{3},\frac{2}{3}]$. Thus although the positive evidence for
necessity

\qed\end{example}

\newpage
